A Test of the Frequency Independence Assumption of Power System Coefficients used in Geomagnetically Induced Current Estimates

### 2019-07-31 paper-submitted-2.tex

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are computed from data and are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system components. It is also shown that further improvements over Model~2 can be made using models with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model. Two additional models are considered that require only the geomagnetic field as an input and they are shown to be significantly better at estimating GIC than the models that use only the geoelectric field as an input.

@@

Data for the time span of 2006-08-04T15:00Z through 2006-08-08T15:00Z are shown Figure~\ref{sample}. The GIC data were provided as 1-day files that start at midnight JST and the times shown in Figure~\ref{sample} are Universal Time.

@@

The GIC data contains non-physical spikes that are followed by 40-80~s of anomalous values that appear as an exponential decay back to pre-spike values. These spikes were removed by identifying times $t$ when the change in the 1~s absolute value of $GIC(t)$ was greater than $0.05$~Amps and then values in a window $GIC(t-2)$ through $GIC(t+100)$ were replaced using a linear interpolation of the values at $GIC(t-3)$ and $GIC(t + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed. For the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

\cite{Lehtinen1985} justifies the use of the quasi-DC approximation using the argument that GIC amplitudes are significant at frequencies that are much lower than the 50~Hz power line frequency. Clarified, this is essentially a statement that the reactance is proportional to frequency and although the reactance is typically $\sim$5 times larger than the resistance at 50~Hz \citep{Purchala2005}, the reactance will be $\sim$10 times smaller or less than the resistance at GIC frequencies of $\sim$1~Hz and lower. Given these arguments, the quasi-DC approximation is justified, and an explanation involving a violation of the assumption of a uniform half-space conducting slab Earth using a conductivity that varies with depth also seems unlikely to explain the results as it would need to account for at least an order of magnitude of reactance. (However, \cite{Boteler1994} noted that for certain systems, the $L/R$ time constant may be large enough to warrant inclusion in DC flow calculations; this topic has more recently been pursued by \cite{Oyedokun2013a}.)

@@

In this discussion, we have attempted to address independently the contribution of spatial variations in the electric field magnitude and direction and spatial variations in the electric field spectra, but generally these variations do not occur independently as they are both caused by horizontal variations in the ground conductivity. Thus, the finding that the angle of the electric field that produces maximum GIC differs significantly from the angle of the 187~kV power line for both Model~1 and Model~2 is likely to be consistent with the frequency dependence of $a$ and $b$ found in Model~2. However, testing this claim would require a full network analysis an knowledge of the ground conductivity throughout the network.

@@

When GIC and magnetic field measurements are available, a model of the form of Model~4 can be used. For this particular data set, we have found that such a model provides the best estimates of GIC in terms of the PE and MSE metrics. Local ground magnetic field estimates along with GIC measurements could be used to derive a transfer function model that is updated when the system configuration changes and this model could be used to estimate GIC under extreme or simulated geomagnetic conditions. A key advantage of this approach is that long-term geomagnetic field measurements are generally easier to make and are more readily available. A disadvantage is the transfer function parameters are difficult to interpret physically because they contain components of the transfer functions that connect the electric field to the magnetic field and GIC to the electric field and frequency-dependent artifacts due to violations of the assumption that the field is spectrally uniform.

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory was obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. We would like to thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Benabeau for review and comments on part (1) of Section\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average have been subtracted from each time series.}
  \label{sample}
\end{figure}

>>>>

\begin{abstract}
  A common assumption used when estimating geomagnetically induced currents (GICs) in a power system given a time series of nearby direct measurements or indirect estimates of the horizontal geoelectric field components $E_x(t)$ and $E_y(t)$ on Earth's surface is that the system is resistive. That is, the approximation $GIC(t)$  $=$ $a_oE_x(t) + b_oE_y(t)$ (Model 1) is used, where $a_o$ and $b_o$ are frequency-independent power system coefficients.  A first test of this assumption is made using GIC measurements in a 187 kV transformer connected to a $\sim$100~km power line in Memanbetsu, Japan and geoelectric field measurements made at the Memanbetsu Magnetic Observatory $\sim$9~km away. A second model (Model 2) is obtained using a frequency domain generalization of Model 1: $GIC(\omega)$ $=$ $a(\omega)E_x(\omega) + b(\omega)E_y(\omega)$. The coefficients $a(\omega)$ and $b(\omega)$ are computed from data and are shown to be frequency dependent and this model provides significantly better estimates of the measured GIC than Model~1. Based on results using a simulated geoelectric field, it is suggested that the measurement-derived frequency dependence of the system coefficients may be explained by spatial variations in the spectrum of the geoelectric field along the line in which GIC is measured rather than by an actual frequency dependence of the system components. It is also shown that further improvements over Model~2 can be made using frequency-dependent models with the geomagnetic field as an input.
\end{abstract}

@@

In this work, we use a unique dataset in which time series measurements of $\mathbf{E}$ and the surface horizontal geomagnetic field, $\mathbf{B}$, were made near a site where GIC was being measured in an electric power system.  We first estimate the system coefficients $a_o$ and $b_o$ from data using conventional methods, in which they are modeled as frequency independent. Then, we use a model in which the system coefficients are allowed to be frequency dependent, compute the frequency dependence, and compare this model's ability to represent GIC with the traditional frequency-independent model. Two additional frequency-dependent models are considered that require only the geomagnetic field as an input and they are shown to be significantly better at estimating GIC than the models that use only the geoelectric field as an input.

@@

Data for the time span of 2006-08-04T15:00Z through 2006-08-08T15:00Z are shown in Figure~\ref{sample}. The GIC data were provided as 1-day files that start at midnight JST and the times shown in Figure~\ref{sample} are Universal Time.

@@

The GIC data contains non-physical spikes that are followed by 40-80~s of anomalous values that appear as an exponential decay back to pre-spike values. These spikes were removed by identifying times $t$ when the change in the 1~s absolute value of $GIC(t)$ was greater than $0.05$~Amps and then values in a window $GIC(t-2)$ through $GIC(t+100)$ were replaced using a linear interpolation of the values at $GIC(t-3)$ and $GIC(t + 101)$. In the example interval shown in Figure~\ref{sample}, 493 spikes were removed and approximately 15 spikes that were were not removed by the automated algorithm are still visible. For the time intervals considered, the fraction of interpolated GIC values ranged from 1--4\%.

@@

\cite{Lehtinen1985} justifies the use of the quasi-DC approximation using the argument that GIC amplitudes are significant at frequencies that are much lower than the 50~Hz power line frequency. Clarified, this is essentially a statement that the reactance is proportional to frequency and although the reactance is typically $\sim$5 times larger than the resistance at 50~Hz \citep{Purchala2005}, the reactance will be $\sim$10 times smaller or less than the resistance at GIC frequencies of $\sim$1~Hz and lower. Given these arguments, the quasi-DC approximation is justified, and an explanation involving a violation of the assumption of a uniform half-space conducting slab Earth using a conductivity that varies with depth also seems unlikely to explain the results as it would need to account for at least an order of magnitude of reactance. (However, \cite{Boteler1994} noted that for certain systems, the $L/R$ time constant of the power network may be large enough to warrant inclusion in GIC flow calculations; this topic has more recently been pursued by \cite{Oyedokun2013a}.)

@@

In the discussion of (3) above, we have attempted to address independently the contribution of spatial variations in the electric field magnitude and direction and spatial variations in the electric field spectra, but generally these variations do not occur independently as they are both caused by horizontal variations in the ground conductivity. Thus, the finding that the angle of the electric field that produces maximum GIC differs significantly from the angle of the 187~kV power line for both Model~1 and Model~2 is likely to be consistent with the frequency dependence of $a$ and $b$ found in Model~2. However, testing this claim would require a full network analysis and knowledge of the ground conductivity throughout the network.

@@

When GIC and magnetic field measurements are available, a model of the form of Model~4 can be used. For this particular data set, we have found that such a model provides the best estimates of GIC in terms of the PE and MSE metrics. Local ground magnetic field estimates along with GIC measurements could be used to derive a transfer function model that is updated when the system configuration changes and this model could be used to estimate GIC under extreme or simulated geomagnetic conditions. A key advantage of this approach is that long-term geomagnetic field measurements are generally easier to make and more readily available than geoelectric field measurements. A disadvantage is the transfer function parameters are difficult to interpret physically because they contain components of the transfer functions that connect the electric field to the magnetic field and GIC to the electric field and frequency-dependent artifacts due to violations of the assumption that the field is spectrally uniform.

\acknowledgments
Magnetic and electric field measurements made at the Kakioka Magnetic Observatory were obtained from the Japan Meteorological Agency Data Portal \url{http://www.kakioka-jma.go.jp/obsdata/metadata/en}. The geomagnetic coordinates in Figure~\ref{map} were computed using code from \url{https://www.ngdc.noaa.gov/geomag/geom_util/apex.shtml}. We would like to thank the Hokkaido Electric Power Co. for the GIC measurements at the Memanbetsu substation and Shinichi Watari for communicating and preparing the GIC data and providing feedback on the manuscript. We also thank Ikuko Fujii for providing clarifications on the geoelectric instrument response, Trevor Gaunt and Emanuel Benabeau for review and comments on part (1) of Section~\ref{discussion}, and Josh Rigler for substantial feedback and review.
%\clearpage

@@

\begin{figure}[h]
  \centering
  \includegraphics[width=\textwidth]{figures/plot_raw_All_20060805-v1.pdf}
  \caption{(Top) Geomagnetic field measured at the Memanbetsu Magnetic Observatory (MMB). (Middle) Geoelectric field measured at the Memanbetsu Magnetic Observatory (MMB). (Bottom) GIC measurements at Memanbetsu substation. The subscripts $x$ and $y$ correspond to Geographic North and East, respectively, the date labels are for Universal Time, and the average has been subtracted from each time series.}
  \label{sample}
\end{figure}
